---
title: "P8105_mtp_bz2561"
author: "Bohan Zhu"
date: "2025-10-18"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy = TRUE)
```

```{r}
library(tidyverse)
library(readxl)
library(patchwork)
```

```{r, include=FALSE}
knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%",
  fig.retina = 2,
  dpi = 320
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```


# Import and Tidy Data
**I assume cleaning rule for FHP size is same as EOP size**
```{r}
exostosis_df = 
  read_excel("data/p8105_mtp_data.xlsx", skip = 8, na = c("NA", ".", "")) |> 
  janitor::clean_names() 
  
exostosis_clean = 
  exostosis_df |> 
  mutate(
    sex= case_match(
      sex, 
      0 ~ "female",
      1 ~ "male"
    ),
    age_group = case_match(
      age_group,
      "2" ~ "18_30",
      "3" ~ "31_40",
      "4" ~ "41_50",
      "5" ~ "51_60",
      c("6", "7", "8") ~ "60+"
    ) |> 
      factor(levels = c("18_30", "31_40", "41_50", "51_60", "60+"),
           ordered = TRUE),
    eop_size_mm = if_else(is.na(eop_size_mm), 0, eop_size_mm),
    eop_size = case_match(
    eop_size,
    "0" ~ "0_5",
    "1" ~ "5_10",
    "2" ~ "10_15",
    "3" ~ "15_20",
    "4" ~ "20_25",
    "5" ~ "25+"
    ) |> 
      factor(levels = c("0_5", "5_10", "10_15", "15_20", "20_25","25+"),
           ordered = TRUE),
    eop_visibility_classification = case_match(
      eop_visibility_classification,
      0 ~ "eop size 0",
      1 ~ "0 < eop size  <=5",
      2 ~ "eop size >= 5"
    ) |> 
      factor(levels = c("eop size 0", "0 < eop size  <=5", "eop size >= 5"),
           ordered = TRUE),
    fhp_size_mm = if_else(is.na(fhp_size_mm), 0, fhp_size_mm),
    fhp_category = case_match(
      fhp_category,
      "0" ~ "0_10",
      "1" ~ "10_20",
      "2" ~ "20_30",
      "3" ~ "30_40",
      "4" ~ "40_50",
      "5" ~ "50_60",
      "6" ~ "60_70",
      "7" ~ "70+"
    ) |> 
      factor(levels = c("0_10", "10_20", "20_30", "30_40", "40_50",
                        "50_60","60_70","70+"),
           ordered = TRUE)
  )
```
The dataset was cleaned by skipping unused header rows, standardizing variable names, and recoding numeric codes into clear labels. `Sex` was relabeled as male/female, and `age groups` as 18–30, 31–40, etc.

Missing `EOP` size and `FHP` size measurements were replaced with zeros, and both `EOP size` and `FHP_category` classifications were converted into readable, ordered factor levels. The variable `eop_visibility_classification` was also recoded to reflect visibility categories.

```{r}
exostosis_clean |> 
  summarise(n())

exostosis_clean |>
  group_by(age_group, sex) |>
  summarise(participants = n(), .groups = "drop") |>
  pivot_wider(names_from = sex, values_from = participants) |>
  knitr::kable()
```
After cleaning, the resulting dataset contained `r nrow(exostosis_clean)` observations and `r ncol(exostosis_clean)` variables. Among the variables, several are key to the analysis. 

`EOP size` and `eop_visibility_classification` capture the prominence of the external protuberance. `FHP size` and `FHP category` are also important, as they measure forward head posture potentially linked to EOP growth. Finally, `sex` and `age group` offer demographic context for biological or behavioral differences.

```{r}
exostosis_clean |> 
  filter(is.na(age_group))

exostosis_df |>
  filter(
    (eop_size_mm >= 0 & eop_size_mm < 5 & eop_size != 0) |
    (eop_size_mm >= 5 & eop_size_mm < 10 & eop_size != 1) |
    (eop_size_mm >= 10 & eop_size_mm < 15 & eop_size != 2) |
    (eop_size_mm >= 15 & eop_size_mm < 20 & eop_size != 3) |
    (eop_size_mm >= 20 & eop_size_mm < 25 & eop_size != 4) |
    (eop_size_mm >= 25 & eop_size != 5)
  ) |>
  select(eop_size_mm, eop_size)|> 
  knitr::kable()

exostosis_clean |> 
  group_by(eop_shape) |> 
  summarise(participants = n()) |> 
  knitr::kable()

exostosis_clean |> 
  filter(is.na(fhp_size_mm))

exostosis_clean |>
  filter(
    (fhp_size_mm >= 0  & fhp_size_mm < 10 & fhp_category != "0_10") |
    (fhp_size_mm >= 10 & fhp_size_mm < 20 & fhp_category != "10_20") |
    (fhp_size_mm >= 20 & fhp_size_mm < 30 & fhp_category != "20_30") |
    (fhp_size_mm >= 30 & fhp_size_mm < 40 & fhp_category != "30_40") |
    (fhp_size_mm >= 40 & fhp_size_mm < 50 & fhp_category != "40_50") |
    (fhp_size_mm >= 50 & fhp_size_mm < 60 & fhp_category != "50_60") |
    (fhp_size_mm >= 60 & fhp_category != "60_70" & fhp_category != "70+")
  ) |>
  select(fhp_size_mm, fhp_category) |> 
  knitr::kable()


exostosis_df |>
  mutate(fhp_category = as.numeric(fhp_category)) |>
  filter(fhp_category > 7) |> 
  select(fhp_category)
```

Some categorical variables do not align with their continuous values. For example, `age_group` has two missing cases—one aged 0–17 (outside defined groups) and another aged 45 misclassified into group 1.

Some `eop_size_mm` and `fhp_size_mm` values are misclassified, some may due to inconsistent boundary rules (unmatchable data listed in table above). 

Besides, `eop_shape` includes categories 1, 2, 3, and 5 but lacks 4, with 500+ NAs suggesting incomplete coding.

One `fhp_category` record (30.8) should fall in the 30–40 range but was uncategorized.

# Visualization

```{r}
Figure_3_advanced = 
  exostosis_clean |>
  filter(!is.na(age_group), !is.na(fhp_size_mm)) |>
  ggplot(aes(x = age_group, y = fhp_size_mm, fill = sex)) +
  geom_violin(width = 0.7, alpha = 0.5) +
  geom_boxplot(width = 0.2, position = position_dodge(width = 0.7)) +
  labs(
    title = "Distribution of FHP by Age Group and Sex",
    x = "Age Group (years)",
    y = "FHP Size (mm)",
    fill = "Sex"
  ) +
  theme(legend.position = "right",
        axis.title.y = element_text(size = 8),
    plot.caption = element_text(size = 5, hjust = 0.5)
  )
```

```{r}
Figure_4_advanced =
  exostosis_clean |>
  mutate(
    enlarged = 
      if_else(eop_size %in% c("10_15", "15_20", "20_25", "25+"), 1, 0)
  ) |>
  filter(!is.na(age_group), !is.na(sex)) |>    
  group_by(age_group, sex) |>
  summarise(
    total = n(),
    enlarged_n = sum(enlarged, na.rm = TRUE),
    rate = enlarged_n / total * 100
  ) |>
  ggplot(aes(x = age_group, y = rate, color = sex, group = sex)) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  scale_y_continuous(labels = scales::percent_format(scale = 1)) +
  labs(
    title = "Rate of Enlarged EOP by Age Group and Sex",
    x = "Age Group (years)",
    y = "Rate of Enlarged EOP",
    color = "Sex"
  ) +
  theme(legend.position = "right",
        axis.title.y = element_text(size = 8),
    plot.caption = element_text(size = 5, hjust = 0.5)
)
```

```{r}
combined_plot = 
  (Figure_3_advanced / Figure_4_advanced) +
   plot_annotation(
    caption = "Combined visualization of FHP and EOP patterns across age and sex groups."
  )
combined_plot
```

Males show greater forward head protraction and higher enlarged EOP prevalence, consistent with the pattern described by the authors. both FHP and EOP measures show a mid-age decline followed by an increase in older groups.

```{r}
exostosis_clean |>
  filter(!is.na(age_group)) |>
  ggplot(aes(x = fhp_size_mm, y = eop_size_mm, color = sex)) +
  geom_point(alpha = 0.7, size = 1.5) +
  facet_grid(sex ~ age_group) +
  scale_color_manual(values = c("female" = "navy", "male" = "green")) +
  labs(
    title = "Association between FHP Size and EOP Size across Age and Sex",
    x = "Forward Head Protraction Size (mm)",
    y = "EOP Size (mm)",
    color = "Sex"
  ) +
  theme_bw(base_size = 11) +
  theme(
    strip.text = element_text(size = 10, face = "bold"),
    plot.title = element_text(hjust = 0.5, face = "bold")
  )
```

The plot shows a weak FHP–EOP relationship across age and sex. Data are scattered with no clear trend. Males show greater variability and larger EOP sizes, suggesting sex seems to have a stronger influence than FHP.

# Reproducing reported results and discussion
```{r}
exostosis_clean |>
  drop_na(age_group) |>
  group_by(age_group) |>
  summarise(participants = n()) |>
  knitr::kable()
```

Small discrepancies exist between the authors’ stated and observed sample sizes, with 21 more participants overall and differences in all age groups except 51–60.

```{r}
exostosis_clean |>
  group_by(sex) |>
  summarise(
    mean_FHP = round(mean(fhp_size_mm), 2),
    sd_FHP = round(sd(fhp_size_mm), 2),
  ) |> 
  knitr::kable()
```
The authors reported mean FHP of $28 \pm 15$ mm in males and $24 \pm 11$ mm in females. Our sample shows $23.65 \pm 10.69$ mm for males and $28.32 \pm 14.80$ mm for females, largely consistent with the authors’ findings.

```{r}
exostosis_clean |>
  summarise(
    EEOP_prevalence = (sum(eop_size_mm > 10, na.rm = TRUE) /
                       sum(!is.na(eop_size_mm))) * 100
  )
```
EEOP defined as an EOP exceeding 10 mm in size. Results based on sample equals $32.1 \% \approx 33\%$ (According to the author).

```{r}
exostosis_clean |>
  drop_na(age_group) |>
  group_by(age_group) |>
  summarise(
    FHP_over_40_percent = (sum(fhp_size_mm > 40) / n()) * 100
  )
```
Among participants over 60, $32.5%$ (close to $34.5%$) had FHP > 40 mm — the highest rate across groups, consistent with both the trend and values reported by the authors.

**Conclusion**

Results largely match the report, though coding and visualization issues reduce clarity. The data show associations, not causation; verifying the phone-use hypothesis needs detailed posture, usage, and longitudinal evidence.
